---
title:
  # Title must be 150 characters or less
  "Active travel and cardiovascular health: A UK-wide ecological study"
csl: bmj.csl
bibliography: Transport-health.bib
---

```{r, echo=FALSE}
# R setup can go here
# # for pdf output add this after the title:
# author:
#   - name: Sarah Walpole
#     affiliation: '1'
#     # If corresponding author (only one corresponding author per manuscript)
#     email: example@example.ac.uk
#   - name: Mark Green
#     affiliation: '2'
#   - name: Robin Lovelace
#     affiliation: '3'
# affiliation:
#   - id: '1'
#     city: London
#     country: UK
#   - id: '2' 
#     city: Liverpool
#     country: UK
#   - id: '3' 
#     city: Leeds
#     country: UK
# output: rticles::plos_article

```


# Abstract {-}

Goes here

# Example headers and content {-}

This section is designed to show off some of RMarkdown's functionality.

## Subsection {-}
You can use R chunks directly to plot graphs. See Figure 1.

```{r, graph, echo=TRUE, message=FALSE, fig.height=4, fig.width=4, fig.align='center', fig.cap='Figure caption'}
require("ggplot2")
x <- 0:100
y <- 2 * (x + rnorm(length(x), sd = 3) + 3)
ggplot(data = data.frame(x, y), 
       aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

Cite fancy references using this notation [@Furie2012]:

```{}
[@Furie2012]
```


# Introduction {-}

What is known:

* Physical activity reduces blood pressure, overweight, and coronary heart
disease (CHD) risk at the individual level
* Cycling vs public transport vs car impacts on local and global environment
* Air pollution exacerbates risk of CHD
* CHD costs the health service and the economy to the tune of 12% of European
and Canadian healthcare costs [@Tarride2009] and 17% of US health expenditure
[@Heidenreich2011]

What this study can add:

* Understanding of whether, at the population level, physical activity achieved
through active transport may be associated with reduced burden of cardiovascular
morbidity and mortality

Aims 

- To explore the relationship between cardiovascular disease and prevalence of
active transport at population level in order to inform policy making
- To explore confounding or moderating factors that may influence this
relationship, including socioeconomic status, smoking prevalence, (gender,
ethnicity and urbanity)

Objectives

- Obtain data on prevalence of active transport, average distance travelled per
person using active transport, under 75 cardiovascular mortality rate,
prevalence of hospital admission for acute coronary syndrome (ACS), median blood
pressure, prevalence of high blood pressure, prevalence of diabetes, median BMI,
prevalence of overweight or obesity, prevalence of  smoking and socioeconomic
status at the unitary authority (UA) and district level in the UK
- Correlate active travel to work (the explanatory variable) with cardiovascular
disease measures (the outcome variable)
- Explore the impact of  potential confounding or  mediating factors:
socioeconomic status, smoking prevalence and prevalence of diabetes, overweight
and obesity
- Explore/validate the correlation between prevalence of active
transport/average distance travelled per person using public transport and
population physical activity levels using the data on “percentage of physically
active adults” from PHE

# Background {-}

Studies from in middle income and high income countries, and in rural and urban
settings have investigated the association between active travel and heart
disease or proxy measures of heart disease risk. Population level studies show
association between higher levels of physical activity and lower levels of
cardiovascular risk. For example, a prospective cohort study of 23000 Dutch
adults used a questionnaire to assess physical activity, and cardiovascular
disease incidence at follow up after an average of 9.8 years. The hazard ratio
of cardiovascular disease was 0.82 (95% CI 0.71-0.95) in those engaged in
cycling, but a dose response relationship was not identified. Walking was not
associated with lower cardiovascular disease hazard ratio at 10 year follow up.
Results were adjusted for age, sex, other physical activities, smoking, alcohol
consumption, and educational level [@Hoevenaar-Blom2011]. This
correlates with findings from *physiological studies* which demonstrate that
physical activity reduces the risk of overweight, hypertension and
cardiovascular risk [REFs, e.g. Goldberg et al.].

A retrospective cohort study used data from a national survey of UK residents
(n=XX)  in 2009-2011 to investigate the association between active transport and
overweight, obesity, hypertension and diabetes in different socio-demographic
groups. This study included public transport, walking and cycling in the
definition of active transport and used logistic regression methods. Across all
groups, active transport was associated with a likelihood of being overweight
(AOR 0.80 (0.54-0.88)) and walking was associated with a lower likelihood of
having hypertension (AOR 0.83 (0.71-0.97)) [@Laverty2013a]. A
retrospective cohort study of US 9933 adults taken from 2007-2008 and 2009-2010
National Health and Nutrition Examination Survey data classified participants by
amount of active transportation and used multivariable linear and logistic
regression analyses to investigate relationship with mean waist circumference
and BMI. The study found a dose-response relationship; participants with low and
high levels of active transport had lower BMI (-0.9 (95% CI -1.4, -0.5) and -1.2
(95% CI -1.7, -0.8) respectively), mean waist circumference (-2.2cm (95% CI
-3.2, -1.2) and -3.1cm (95% CI -4.3, -1.9)) and likelihood of hypertension
(AOR=0.76 (95% CI 0.61, 0.94) and AOR=0.69 (95% CI 0.58, 0.83)) compared to
those reporting no active transport. High active transport level was associated
lower likelihood of being diabetic (AOR=0.69 (95% CI 0.54, 0.88)), but no
association was seen with high-density lipoprotein level. The study controlled
for sociodemographic characteristics, smoking status, and minutes/week of
physical activity not related to transport [@Furie2012].

A retrospective cardiovascular study investigated the relationship between
active travel to neighbourhood amenities (e.g. parks) and BMI, waist
circumference and cardiovascular fitness in 3547 participants. Analysis was with
multivariable linear or multinomial regression models, stratified by gender.
More participants walked than cycled to neighbourhood amenities, factors such as
having a partner or having children were associated with lower likelihood of
using active transport. Walking or cycling were associated with more favorable
BMI (-1.68 (95%CI -2.81, -0.55), waist circumference (-3.41cm (95%CI -5.71,
-1.11)) and treadmill test duration (36.65seconds (95%CI (17.99, 55.31)).
[@Boone-Heinonen2009]

In middle income countries, similar results are seen. A retrospective cohort
study of 8400 adults in China supports the association between lower likelihood
of obesity and walking or cycling for transport, but found that walking or
cycling was associated with increased likelihood of having diabetes or high
cholesterol. This study collected travel data using participant interviews, and
carried out physiological and biochemical tests. [@Lu2013].  A
retrospective cohort study of women in China found inverse relationships between
cycling for transport and all-cause mortality and walking for transport and
all-cause mortality, although the latter was not significant (p=0.07)
[@Matthews2007]. A cross-sectional study of 1,366 rural and 2,536 urban
participants
in India used random-effect logistic regression models adjusted for age, sex,
caste, standard of living, occupation, factory location, leisure time physical
activity, daily fat intake, smoking status, and alcohol use, and found adjusted
risk ratios of for overweight/obesity of 0.72 (95% CI 0.58-0.88) and 0.66 (95%
CI 0.55-0.77) for walking and cycling to work respectively, compared to using
private motor transport. Cycling to work compared to using private transport was
associated with an adjusted risk ration of 0.51 (95% CI 0.36-0.71) for
hypertension and 0.65 (95% CI 0.44-0.95) for diabetes [@Millett2013]. Like
the USA study by @Furie2012, this study provides evidence of a dose
response relationship between active transport and reduction in cardiovascular
risk.

There is evidence that the benefits of active travel may be more beneficial to
some socio-demographic groups than others. For example, an RCT involving
overweight men (n=60) and women (n=71) all aged 61-79 years found that about 40
minutes of cycling, an average of 2.5 times per week resulted in an average 10.6
and 5.5 mmHg reduction in blood pressure, and XX and XX reduction in body mass
index in the groups respectively [@Morita2013]. At the population
level active travel is found to have more health benefits in urban residents
than rural residents [@Millett2013]. [@Laverty2013a] – more uptake of
walking in black ethnicity participants) A cross sectional study of 344 10-year
old children in USA at two time periods, six years apart suggests that the
association between active transport and cardiovascular risk holds seen in
adults is also present in children [@Andersen2011].

# Method {-}

Data sources

Cycling data from the Transport survey for England or the UK Census
Cardiovascular data from Public Health England (PHE) or Farr Institute data
PHE data is by strategic clinical network and includes: 

- CHD prevalence (QOF)
- CHD mortality in under 75s (age standardised)
- (Heart failure, various parameters)
- obesity
- prevalence low physical activity (?correlate this with cycling?)
- prevalence hypertension
- percentage of people in most deprived quintile (20%?!)

Statistical analysis
Programmes: R,
Analysis

# Results {-}

Results and Discussion can be combined.

# Discussion {-}

Findings

Implications 

- correlation, not causation, mediating factors – individual and population level
- individual vs societal risks/benefits

Exploration of robustness

- Methods
- data sources

Generalisability and transferability of findings
Recommendations/ Implications for policy makers

- primary vs secondary prevention

# Conclusion {-}

# Acknowledgments {-}

# References {-}

